---
title: "Getting Started with exportr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with exportr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

```{r setup}
library(exportr)
library(data.table)  # For fwrite_with_metadata
```

## Introduction

The `exportr` package enhances standard R data export functions by automatically tracking data provenance. Every time you export data, `exportr` records which R script created the file, making it easy to trace data files back to their source code.

## Why Use exportr?

### The Problem
In research and data analysis workflows, it's common to:

- Generate multiple output files from different scripts
- Lose track of which script created which file  
- Need to recreate analyses but forget the source code
- Share data files without documentation of how they were created
- Struggle with data lineage and reproducibility

### The Solution
`exportr` solves these problems by automatically embedding metadata about the source script directly into the exported files. **The most important piece of metadata is the full path to the R script that created the file.**

## How It Works

1. **Save the file normally**: Your data is saved exactly as it would be with the standard function
2. **Detect the source script**: `exportr` automatically identifies which R script is calling the function  
3. **Add metadata**: The script path and other information is stored alongside the file

**Key Point**: The exported data files contain exactly the same data as the original functions would produce - no modifications to file contents!

## Basic Usage

Load the package and use the enhanced export functions exactly like you would use the standard ones:

```{r basic_usage}
library(exportr)

# Standard approach:
# fwrite(iris, "iris.csv")

# Enhanced approach with automatic metadata:
fwrite_with_metadata(iris, "iris.csv")

# Check what metadata was added
verify_file_metadata("iris.csv")
```

## Available Functions

### Enhanced Export Functions

All functions work exactly like their standard counterparts but add automatic metadata:

```{r functions}
# Enhanced data.table::fwrite() - PRIORITY: script path tracking
fwrite_with_metadata(mtcars, "cars.csv")
read_exportr_metadata("cars.csv")

# Enhanced utils::write.csv()
write_csv_with_metadata(iris, "flowers.csv")

# Enhanced base::saveRDS()
model <- lm(mpg ~ wt + hp, data = mtcars)
saveRDS_with_metadata(model, "car_model.rds")
```

### Utility Functions

```{r utilities}
# Check if a file has exportr metadata
has_exportr_metadata("cars.csv")

has_exportr_metadata("car_model.rds")
metadata <- read_exportr_metadata("car_model.rds")
print(metadata)
# Read metadata from a file
metadata <- read_exportr_metadata("cars.csv")
print(metadata$script_path)

# Verify metadata (detailed view)
verify_file_metadata("cars.csv")

# Get current script information
script_info <- get_script_info()
print(script_info)
```

## Understanding Metadata Storage

Unlike image files that support EXIF data, CSV and RDS files need different approaches for metadata storage.

### Where Metadata Is Stored

#### Windows
- **Primary method**: Alternate data streams
- **Location**: `filename.csv:exportr.script_path`
- **Verification**: Command Prompt: `dir /R filename.csv`
- **Access**: `more < "filename.csv:exportr.script_path"`

#### macOS  
- **Primary method**: Extended attributes via `xattr`
- **Location**: `com.exportr.script_path` attribute
- **Verification**: Terminal: `xattr -l filename.csv`
- **Access**: `xattr -p com.exportr.script_path filename.csv`

#### Linux
- **Primary method**: Extended attributes via `setfattr`
- **Location**: `user.exportr.script_path` attribute  
- **Verification**: Terminal: `getfattr -d filename.csv`
- **Access**: `getfattr -n user.exportr.script_path filename.csv`

#### All Platforms (Fallback)
- **Fallback method**: Hidden sidecar files
- **Location**: `.filename.csv.exportr` files alongside your data
- **Contents**: JSON or human-readable metadata

## Metadata Modes

You can control how metadata is stored using the `metadata_mode` parameter:

```{r metadata_modes}
# Default: Try file attributes first, fallback to sidecar
fwrite_with_metadata(iris, "iris_default.csv", metadata_mode = "attribute")

# Force sidecar file creation
fwrite_with_metadata(mtcars, "cars_sidecar.csv", metadata_mode = "sidecar")

# Create log file (CSV format for easy analysis)
fwrite_with_metadata(trees, "trees_log.csv", metadata_mode = "log")

# No metadata tracking
fwrite_with_metadata(airquality, "air_none.csv", metadata_mode = "none")
```

## Real-World Example: Research Workflow

Let's simulate a typical research workflow to see exportr in action:

```{r workflow_example}
# =============================================================================
# Step 1: Data Preparation Script (save as "data_prep.R")
# =============================================================================

# Load and clean raw data
raw_data <- data.frame(
  id = 1:1000,
  group = sample(c("Control", "Treatment A", "Treatment B"), 1000, replace = TRUE),
  baseline = rnorm(1000, mean = 100, sd = 15),
  followup = rnorm(1000, mean = 105, sd = 18)
)

# Clean the data
clean_data <- raw_data[complete.cases(raw_data), ]
clean_data$change_score <- clean_data$followup - clean_data$baseline

# Export cleaned data with metadata
fwrite_with_metadata(clean_data, "data/cleaned_dataset.csv")

# =============================================================================
# Step 2: Analysis Script (save as "statistical_analysis.R") 
# =============================================================================

# Read the cleaned data
clean_data <- fread("data/cleaned_dataset.csv")

# Perform statistical analysis
library(dplyr)
results <- clean_data %>%
  group_by(group) %>%
  summarise(
    n = n(),
    mean_baseline = mean(baseline),
    mean_followup = mean(followup),
    mean_change = mean(change_score),
    se_change = sd(change_score) / sqrt(n()),
    .groups = 'drop'
  )

# Export results with metadata
fwrite_with_metadata(results, "results/group_analysis.csv")

# Create and save statistical model
model <- lm(change_score ~ group, data = clean_data)
saveRDS_with_metadata(model, "models/change_model.rds")

# =============================================================================
# Step 3: Later - Check File Provenance
# =============================================================================

# Check where each file came from
verify_file_metadata("data/cleaned_dataset.csv")
verify_file_metadata("results/group_analysis.csv") 
verify_file_metadata("models/change_model.rds")
```

Expected output:
```
Metadata for cleaned_dataset.csv:
  Script path: C:/project/scripts/data_prep.R
  Created: 2025-07-25 14:30:45
  User: researcher
  Detection method: rstudio_api

Metadata for group_analysis.csv:
  Script path: C:/project/scripts/statistical_analysis.R
  Created: 2025-07-25 15:15:22
  User: researcher
  Detection method: rstudio_api

Metadata for change_model.rds:
  Script path: C:/project/scripts/statistical_analysis.R
  Created: 2025-07-25 15:15:30
  User: researcher
  Detection method: rstudio_api
```

## Integration with Existing Code

`exportr` functions are designed as drop-in replacements. To add metadata tracking to existing code:

```{r integration}
# Before: Standard R functions
library(data.table)
fwrite(my_data, "output.csv")
write.csv(summary_stats, "summary.csv")
saveRDS(fitted_model, "model.rds")

# After: Enhanced functions with metadata tracking
library(exportr)
fwrite_with_metadata(my_data, "output.csv")
write_csv_with_metadata(summary_stats, "summary.csv")  
saveRDS_with_metadata(fitted_model, "model.rds")
```

## Verification and Troubleshooting

### Check System Capabilities

```{r system_check}
# Check what metadata storage methods are available on your system
system_info()

# Test script detection
test_script_detection()
```

### Verify Metadata Was Written

```{r verification}
# Method 1: Use exportr's verification function
verify_file_metadata("my_file.csv")

# Method 2: Check if metadata exists
has_exportr_metadata("my_file.csv")

# Method 3: Read metadata back
metadata <- read_exportr_metadata("my_file.csv")
print(metadata)
```

### Platform-Specific Verification

#### Windows (Command Prompt)
```cmd
# Check for alternate data streams
dir /R "my_file.csv"

# Read script path
more < "my_file.csv:exportr.script_path"
```

#### macOS (Terminal)
```bash
# Check for extended attributes
xattr -l my_file.csv

# Read script path
xattr -p com.exportr.script_path my_file.csv
```

#### Linux (Terminal)
```bash
# Check for extended attributes
getfattr -d my_file.csv

# Read script path  
getfattr -n user.exportr.script_path my_file.csv
```

### Common Issues and Solutions

#### Script Detection Not Working
```{r troubleshooting_script}
# Check what exportr detects
script_info <- get_script_info()
print(script_info$detection_method)
print(script_info$script_path)

# If detection fails, the method will be "unknown"
# This often happens in interactive sessions
```

#### Metadata Not Being Stored
```{r troubleshooting_metadata}
# Check if metadata was added (should return TRUE)
has_exportr_metadata("your_file.csv")

# Look for sidecar files (fallback method)
list.files(pattern = "\\.exportr$", all.files = TRUE)

# Check system capabilities
caps <- check_system_capabilities()
print(caps$recommended_mode)
```

## Best Practices

### 1. Use Meaningful File Names
Since the script path is tracked, use descriptive output file names:
```{r best_practices_names}
# Good
fwrite_with_metadata(cleaned_data, "results/experiment1_cleaned_data.csv")
fwrite_with_metadata(summary_stats, "results/experiment1_summary_statistics.csv")

# Avoid generic names
# fwrite_with_metadata(data, "output.csv")
# fwrite_with_metadata(results, "temp.csv")
```

### 2. Organize Your Scripts
Keep analysis scripts in logical directories for better traceability:
```
project/
├── scripts/
│   ├── 01_data_preparation.R
│   ├── 02_exploratory_analysis.R
│   ├── 03_statistical_modeling.R
│   └── 04_visualization.R
├── data/
│   ├── raw/
│   └── processed/
└── results/
    ├── figures/
    ├── tables/
    └── models/
```

### 3. Check Metadata Periodically
```{r best_practices_check}
# Verify tracking is working
verify_file_metadata("important_results.csv")

# Check multiple files at once
files_to_check <- c("data.csv", "results.csv", "model.rds")
for (file in files_to_check) {
  if (file.exists(file)) {
    cat("\n")
    verify_file_metadata(file)
  }
}
```

### 4. Document Your Workflow
Combine `exportr` with good commenting and documentation practices:
```{r best_practices_document}
# ==============================================================================
# Data Analysis Pipeline - Step 2: Statistical Modeling
# Author: Researcher Name
# Date: 2025-07-25
# Purpose: Fit linear mixed-effects models to experimental data
# ==============================================================================

# Load preprocessed data (created by 01_data_preparation.R)
analysis_data <- fread("data/processed/experimental_data.csv")

# Fit primary model
primary_model <- lmer(response ~ treatment + (1|subject), data = analysis_data)

# Export model with automatic script path tracking
saveRDS_with_metadata(primary_model, "models/primary_lmer_model.rds")

# Export model results table
model_results <- broom.mixed::tidy(primary_model)
fwrite_with_metadata(model_results, "results/primary_model_coefficients.csv")
```

## Advanced Features

### Custom Metadata Modes
```{r advanced_modes}
# Force specific storage method
fwrite_with_metadata(data, "file.csv", metadata_mode = "sidecar")

# Create both sidecar and log files
fwrite_with_metadata(data, "file.csv", metadata_mode = "sidecar", create_log = TRUE)
```

### Reading Comprehensive Metadata
```{r advanced_reading}
# Read all metadata from all sources
all_metadata <- read_all_metadata("my_file.csv")
print(names(all_metadata))  # Shows: sidecar, attributes, log, etc.

# Compare metadata across different storage methods
compare_metadata_sources("my_file.csv")
```

### Cleaning Up Metadata
```{r advanced_cleanup}
# Remove all metadata from a file
clean_metadata("test_file.csv", 
               remove_sidecar = TRUE,
               remove_attributes = TRUE, 
               remove_metadata_file = TRUE)
```

## Package Configuration

### Set Global Options
```{r configuration}
# Set package-wide defaults
set_exportr_options(
  default_metadata_mode = "attribute",
  verbose = TRUE,
  fallback_to_sidecar = TRUE
)

# Check current options
get_exportr_options()
```

## Conclusion

`exportr` provides a simple but powerful way to maintain data provenance in your R workflows. By automatically tracking which script created each output file, it helps ensure reproducibility and makes collaboration easier.

### Key Benefits:
- **Automatic provenance tracking**: No extra work required
- **Script path priority**: Focus on the most important metadata
- **Drop-in replacements**: Works with existing code
- **Cross-platform**: Windows, macOS, and Linux support
- **Multiple storage methods**: Robust fallback options
- **No file modification**: Original data integrity preserved

### Getting Started:
1. Replace your existing export functions with `_with_metadata` versions
2. Use `verify_file_metadata()` to check that tracking is working
3. Incorporate into your research workflow for better reproducibility

Start using `exportr` today to never lose track of your data provenance again!

## Further Reading

- Check out the package documentation: `help(package = "exportr")`
- Test script detection: `test_script_detection()`
- Verify system capabilities: `system_info()`
- Try the demo: `demo_exportr()`
